---
title: " U.S. National Oceanic and Atmospheric Administration's (NOAA) - Data analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data processing

First is to read the data from the file
```{r readdata}
# Load the data
setwd ("C:/Users/PakHang/Desktop/Coursera JHU/Chapter 5/Data")
StormData <- read.csv("StormData.csv.bz2")

#To start with, we will go through the data 
dim (StormData)
head (StormData, 10)
```
In the data, we can find that there are 692288 observations and 37 variables. However, most hte variable are not relevant to this data analysis topic, so we have to subset the data with the relevant data i.e EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP,CROPDMG, CROPDMGEXP

```{r subsetdata}
stormdatav2<- StormData[c("EVTYPE", "FATALITIES", "INJURIES", "PROPDMG", "PROPDMGEXP","CROPDMG", "CROPDMGEXP")]
```

## Convert the variables 
In the data set, some of the data is not appropriate. Therfore we have to clean up the data before computing.

1.Change the number of fatalities and injuries from factor variables to numeric variables
```{r changenumbers}
stormdatav2$FATALITIES <- as.numeric(stormdatav2$FATALITIES)
stormdatav2$INJURIES <- as.numeric(stormdatav2$INJURIES)
stormdatav2$PROPDMG <- as.numeric(stormdatav2$PROPDMG)
stormdatav2$CROPDMG <- as.numeric(stormdatav2$CROPDMG)
```

2.The Property damange and Compr damange are presented as a form of base and the unit (K= 1000, M= 1,000,000, B = 1,000,000,000,0). Hence we need to 1.Convert the unit in column with PROPDMGEXP and CROPDMGEXP to number and multiply the numerical value by the unit 
```{r set new variables}
#Set a new column to store the numerical value of the unit
stormdatav2["PROPLOSSUNIT"] <- 1
stormdatav2["PROPLOSSVALUE"] <- 1
stormdatav2["CROPLOSSUNIT"] <- 1
stormdatav2["CROPLOSSVALUE"] <- 1

#Convert the units for property damages into integer, and calculate the total loss
if (stormdatav2$PROPDMGEXP == "k " || stormdatav2$PROPDMGEXP == "K") {
   stormdatav2$PROPLOSSUNIT <- 1000
   } else if  (stormdatav2$PROPDMGEXP == "m" || stormdatav2$PROPDMGEXP == "M" ) {
   stormdatav2$PROPLOSSUNIT <- 1000000
   } else if (stormdatav2$PROPDMGEXP == "b"  || stormdatav2$PROPDMGEXP == "B" ) {
   stormdatav2$PROPLOSSUNIT <- 1000000000
   } else {
   stormdatav2$PROPLOSSUNIT <- 1 
   }

stormdatav2["PROPLOSSVALUE"] <- stormdatav2$PROPLOSSUNIT*stormdatav2$PROPDMG

#Convert the units for crop damages into integer, and calculate the total loss
if (stormdatav2$CROPDMGEXP == "k " || stormdatav2$CROPDMGEXP == "K") {
   stormdatav2$CROPLOSSUNIT <- 1000
   } else if  (stormdatav2$CROPDMGEXP == "m" || stormdatav2$CROPDMGEXP == "M" ) {
   stormdatav2$PROPLOSSUNIT <- 1000000
   } else if (stormdatav2$CROPDMGEXP == "b"  || stormdatav2$CROPDMGEXP == "B" ) {
   stormdatav2$CROPLOSSUNIT <- 1000000000
   } else {
   stormdatav2$CROPLOSSUNIT <- 1 
   }
stormdatav2["CROPLOSSVALUE"] <- stormdatav2$CROPLOSSUNIT*stormdatav2$CROPDMG
```

## Compute the personal health statistics
We can use the aggregate function to get the sum of fatalities and injuries in different hazards
```{r aggregate}
stormfatal <- aggregate(FATALITIES ~ EVTYPE, stormdatav2, FUN = sum)
storminj <- aggregate(INJURIES ~ EVTYPE, stormdatav2, FUN = sum)
```
Then we can take a look for the disasters which cause the greatest no.of fatalities 
```{r fatalities}
# Rank the natural hazards according to no.of fatalities (From most to least)
moststormfatal = stormfatal[order(-stormfatal$FATALITIES), ] 
# Take 10 natural hazards which cause the most no.of fatalities 
fataldisaster <- head(moststormfatal, 10)
# Generate the plot to illustrate the statistics
barplot(fataldisaster$FATALITIES, las=3, names.arg = fataldisaster$EVTYPE, main = "Disaster with highest no.of death", ylab = "No.of deaths", col=topo.colors(10))
```

Besides fatalities, the no.of injuries are as follows:
```{r injuries}
# Rank the natural hazards according to no.of injuries (From most to least)
moststorminj = storminj[order(-storminj$INJURIES), ]
# Take 10 natural hazards which cause the most no.of injuries
injdisaster <- head(moststorminj, 10)
# Generate the plot to illustrate the statistics
barplot(injdisaster$INJURIES, las=3, names.arg = injdisaster$EVTYPE, main = "Disaster with highest no.of injuries", ylab = "No.of injuries", col=topo.colors(10))
```

With the statistics, we can see that the STM Wind caused the largest number of death among all the natural hazards.

## Compute the Economical loss
We can use the aggregate function to get the sum of property loss and crop loss in different hazards
```{r proplossaggregate}
propdamage <- aggregate(PROPLOSSVALUE ~ EVTYPE, stormdatav2, FUN = sum)
cropdamage <- aggregate(CROPLOSSVALUE ~ EVTYPE, stormdatav2, FUN = sum)
```
Then we can take a look for the disasters which cause the greatest no.of property loss
```{r proploss}
# Rank the natural hazards according to no.of property loss (From most to least)
propdamage = propdamage[order(-propdamage$PROPLOSSVALUE), ] 
# Take 4 natural hazards which cause the most no.of property loss
toppropdamage <- head(propdamage, 4)
# Generate the plot to illustrate the statistics
barplot(toppropdamage$PROPLOSSVALUE/10^6, las = 3, names.arg = toppropdamage$EVTYPE,main = "Disaster with highest property loss", ylab = "Loss in Million", col=topo.colors(4))
```

Besides fatalities, the no.of injuries are as follows:
```{r croploss}
# Rank the natural hazards according to no.of crop loss (From most to least)
cropdamage = cropdamage[order(-cropdamage$CROPLOSSVALUE), ] 
# Take 4 natural hazards which cause the most no.of crop loss
topcropdamage <- head(cropdamage, 4)
# Generate the plot to illustrate the statistics
barplot(topcropdamage$CROPLOSSVALUE/10^6, las = 3, names.arg = topcropdamage$EVTYPE, main = "Disaster with highest crop loss", ylab = "Loss in Million", col=topo.colors(4))
```

With the statistics, we can see that the STM Wind also caused the greatest economic loss among all the natural hazards.Therefore we can colclude that STM wind cause the most personal and economical damage.
